---
title: "Gapminder_Explore"
author: "Shirlett"
date: "September 26, 2017"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=10)
```

## Bring the Data In
```{r}
library(gapminder)
library(tidyverse)
```

## Smell Test The Data
```{r}

#Find the structure of the object
str(gapminder)
#alternative ways to view the extent and size of the gapminder object ie., summary and #dim
summary(gapminder)
dim(gapminder)

```

Based on the values returned by R's structure command, the gapminder object is a data frame with class table dataframe. There are six variables and 1704 observations. The structure command provides an abbreviated description of an object in R including the number of observations and variables.The summary function provides the information about extent and size but also provides basic statistical information. It is useful to double check results from other statistical functions. The dim function also shows the number of rows x number of columns. This function provides the least amount of information and is useful for determining whether the entire content of the dataframe can be easily viewed on a screen or downloaded to another framework.
The structure function also provides the data type for each variable. Country and Continent are factors or nominal values; year and population are integers; lifeExp and GDP per Capita are numbers.



## Explore Individual Variables
```{r}
#Exploring Continent as a categorical variable
#Tabular description of the spread of Continent
with(gapminder, table(continent))
#bar plot showing the relative frequency of the occurrence of each variable
ggplot(gapminder, aes(x = continent)) + geom_bar(fill = "#FF6666") + ggtitle("Bar Chart showing the Frequency of Occurrence of each Continent")



#Exploring Population as a Quantitative variable
#One line description of the central tendency of Population
summary(gapminder$pop)   
#more detailed description of the population
library(psych)
describe(gapminder$pop)
#density plot to illustrate the spread of population
library(scales)
ggplot(gapminder, aes(x = pop)) + geom_density() + ggtitle("Density Plot showing the Spread of the Population in all Observations") +
scale_x_continuous(labels = comma)


```

For continent, there are six possible values, with Africa occurring most frequently at 624. Oceania occurs the least frequently with 24 observations. Population has a much wider range, with 60,010 as the lowest number and 1,319,000,000 as the highest observation. The average population is 29,600,000 although the value in the middle is 7,024,000. Overall, based on the density plot, the population data is skewed to the right with most values being less than the mean.

## Explore Various Plot Types
### Barchart and Density Plot
```{r}
#This is a basic barchart that shows the spread of population by continent. 
#The library scales was used to format te y-axis. It appears that Asia has the highest #population and will be explored further.
library(scales)
ggplot(gapminder, aes(x = continent, y = pop, color=continent)) + geom_col() +  ggtitle("Bar Chart of the Population in all Continents") + scale_y_continuous(labels = comma)

```
```{r}
#These commands create a variable that shows the countries in Asia and the average #population across the years
library(dplyr)
population <- gapminder %>% 
    filter(continent == "Asia") %>% 
    group_by(country) %>% 
    select(country, pop) %>%
    summarize(avgpop = mean(pop))

#This command creates a density plot to show average population and reorients
#the labels on the x-axis. It shows that China and India have the highest population #levels
ggplot(population, aes(x = country, y = avgpop)) + geom_density() + theme(axis.text.x = element_text(angle = 90, vjust= 0, hjust = 1)) +  scale_y_continuous(labels = comma) + ggtitle("Density Plot of the Avg Population in Asian Countries across all Years") + ylab("Average Population")
```

### ScatterPlot
```{r}
#This command modifies the data so that the GDP per Capita in the most population dense #countries in Asia can be viewed. A new column is added that rounds the GDP per capita.
pop_growth <- gapminder %>% 
    filter(country == "China"| country == "India") %>% 
    mutate(roundgdp = round(gdpPercap,  digits = 0)) %>% 
    select(country, year, pop, roundgdp)


#These commands create a plot that shows population growth in China and India and the 
#associated gdp per Capita. Although the populations of both countries have steadily
#increased since 1952, after 1977, China has had a far greater gdp per capita than #India.
ggplot(pop_growth, aes(x = year, y = pop, color=country, size = roundgdp)) + ggtitle("Plot of the Population Growth with GDP per Capita in China and India") + geom_point() +
geom_text(aes(label=roundgdp, hjust=0.5, vjust=1.5)) + scale_x_continuous(breaks=seq(1952, 2007, 5)) + scale_y_continuous(labels = comma) + ylab("Population")


```


## Extra
### Incorrect Filter Method
```{r}
#Given method of getting Rwanda and Afghanistan whih produces only 12 results
filter(gapminder, country == c("Rwanda", "Afghanistan"))
 

```
### Correct Filter Method
```{r}
#Correct method to obtain Rwanda and Afghanistan which produces all 24 possible results
filter(gapminder, country %in% c("Rwanda", "Afghanistan"))
```

